\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs,longtable,enumitem,hyperref}
\usepackage[T1]{fontenc}

\title{Morphological Profiling to Gene Programs and Gene Sets}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Overview}
This note defines a morphology-based extractor for converting Cell Painting or other imaging-derived morphological profiles into gene-level weights and gene sets. The central use case is a \emph{query profile} measured in a user experiment, which is compared against a \emph{reference library} of profiles from perturbations with semantic handles (compounds, CRISPR knockouts, ORF overexpression). The reference library provides a route from profile similarity to gene hypotheses, and finally to ranked gene sets.

The intended implementation follows the same design principles as other extractors in the repository:
\begin{itemize}[leftmargin=2em]
\item a clear separation between raw assay processing and extractor logic,
\item conservative defaults that produce interpretable outputs,
\item optional resource bundles for reusable references,
\item and explicit metadata, warnings, and GMT output.
\end{itemize}

\section{Biological rationale}
Morphological profiling captures downstream consequences of perturbing cellular systems. A query profile can therefore be interpreted by asking which \emph{known perturbations} in a reference library induce similar morphologies. If the similar reference perturbations are genetic (CRISPR or ORF), the mapping to genes is direct. If they are compounds, the mapping is mediated by a compound-to-target table.

This yields a natural posterior-like gene scoring model:
\begin{enumerate}[leftmargin=2em]
\item compare a query morphology profile to reference perturbation profiles,
\item convert similarity to a perturbation evidence weight,
\item route perturbation evidence to genes through direct or target-based mappings,
\item aggregate evidence across perturbations and modalities,
\item convert the resulting gene weights to one or more gene sets.
\end{enumerate}

In practice, Cell Painting references such as JUMP are especially attractive because they contain large collections of compounds and genetic perturbations with strong metadata, including matched compound/target and gene perturbation panels. Public JUMP resources include a pilot set of 300+ compounds and 160+ genes in A549 and U2OS, a very large production set with over 116,000 compounds and 15,000+ genes in U2OS, and a curated JUMP-Target metadata resource linking compounds, ORFs, and CRISPR reagents to target genes. These resources are publicly documented by the JUMP consortium and the Cell Painting Gallery. \cite{jump_main,jump_gallery,jump_target,jump_pilot,jump_morphmap}

\section{Inputs and notation}
Let
\begin{itemize}[leftmargin=2em]
\item $f \in \{1,\dots,F\}$ index morphology features,
\item $q \in \{1,\dots,Q\}$ index query profiles or query groups,
\item $r \in \{1,\dots,R\}$ index reference perturbation profiles,
\item $g \in \{1,\dots,G\}$ index genes.
\end{itemize}

The query profile matrix is $X^{(Q)} \in \mathbb{R}^{Q \times F}$.
The reference profile matrix is $X^{(R)} \in \mathbb{R}^{R \times F}$.
The extractor assumes these profiles are already computed and aggregated at the appropriate level (for example, well-level normalized and feature-selected profiles from a CellProfiler or pycytominer workflow).

Reference metadata provides:
\begin{itemize}[leftmargin=2em]
\item perturbation type $t_r \in \{\mathrm{compound}, \mathrm{crispr}, \mathrm{orf}\}$,
\item a perturbation identifier $p_r$,
\item optional QC weight $c_r \in [0,1]$,
\item optional cell line, timepoint, and batch labels.
\end{itemize}

Compound target metadata provides a sparse mapping $A \in \mathbb{R}_{\ge 0}^{R \times G}$ where $A_{rg}$ is the weight by which compound perturbation $r$ implicates gene $g$. For genetic perturbations, a direct one-hot mapping $B \in \mathbb{R}_{\ge 0}^{R \times G}$ is used:
\[
B_{rg} =
\begin{cases}
1 & \text{if perturbation } r \text{ directly perturbs gene } g, \\
0 & \text{otherwise.}
\end{cases}
\]

\section{Feature alignment and standardization}
\subsection{Feature intersection}
Let $\mathcal{F}$ be the intersection of features present in the query and reference profiles. Only these features are used. The extractor should warn if the fraction of reference features recovered in the query is low.

\subsection{Reference-based scaling}
If the reference bundle provides feature means and scales (or robust medians and MADs), define standardized query and reference values:
\[
\tilde{X}^{(Q)}_{qf} = \frac{X^{(Q)}_{qf} - \mu_f}{\sigma_f + \epsilon}, \qquad
\tilde{X}^{(R)}_{rf} = \frac{X^{(R)}_{rf} - \mu_f}{\sigma_f + \epsilon}.
\]
If no explicit statistics are supplied, the implementation may assume the profiles are already normalized in a common feature space and leave them unchanged.

\section{Query aggregation}
If multiple query rows correspond to the same biological unit (for example, replicate wells or a user-supplied group), aggregate them before similarity calculation. A robust default is the componentwise median:
\[
\bar{X}^{(Q)}_{uf} = \mathrm{median}\{ \tilde{X}^{(Q)}_{qf} : q \in \mathcal{Q}(u) \},
\]
where $\mathcal{Q}(u)$ is the set of query profiles assigned to unit $u$.
This produces one extracted program per query unit $u$.

\section{Similarity model}
\subsection{Base similarity}
The default similarity between query unit $u$ and reference perturbation $r$ is cosine similarity:
\[
\mathrm{sim}(u,r) = \frac{\bar{X}^{(Q)}_u \cdot \tilde{X}^{(R)}_r}{\|\bar{X}^{(Q)}_u\|_2 \, \|\tilde{X}^{(R)}_r\|_2 + \epsilon}.
\]
An optional alternative is Pearson correlation. The implementation should support both.

\subsection{Polarity}
Morphological similarity can be interpreted in at least two ways:
\begin{itemize}[leftmargin=2em]
\item \textbf{similar} (default): retrieve perturbations with similar morphology,
\item \textbf{opposite}: retrieve perturbations with anti-correlated morphology,
\item \textbf{both}: emit both directions as separate programs.
\end{itemize}

Define positive and negative evidence:
\[
s^+_{ur} = \max(\mathrm{sim}(u,r), 0)^\gamma, \qquad
s^-_{ur} = \max(-\mathrm{sim}(u,r), 0)^\gamma,
\]
where $\gamma \ge 1$ is a similarity sharpness parameter.

\subsection{Perturbation QC weights}
If a perturbation-level QC weight $c_r$ is provided (for example, replicate consistency or profile activity), define:
\[
w^{+}_{ur} = c_r s^+_{ur}, \qquad
w^{-}_{ur} = c_r s^-_{ur}.
\]
If no QC weights are supplied, set $c_r = 1$.

\section{Routing perturbation evidence to genes}
\subsection{Compound perturbations}
For compounds, evidence is distributed across targets according to the compound target map:
\[
C_{rg} = A_{rg}.
\]
If only an unweighted target list is available, use uniform weights within each compound:
\[
\sum_g A_{rg} = 1 \quad \text{for each compound } r \text{ with at least one target.}
\]

\subsection{Genetic perturbations}
For genetic perturbations, evidence maps directly to the perturbed gene:
\[
G_{rg} = B_{rg}.
\]

\subsection{Modality-specific gene scores}
For query unit $u$, define the positive similarity gene evidence from compounds and genetics separately:
\[
E^{+,\mathrm{compound}}_{ug} = \sum_{r : t_r = \mathrm{compound}} w^+_{ur} C_{rg},
\]
\[
E^{+,\mathrm{genetic}}_{ug} = \sum_{r : t_r \in \{\mathrm{crispr}, \mathrm{orf}\}} w^+_{ur} G_{rg}.
\]

To prevent one modality from dominating simply because it has more perturbations, normalize each by its total evidence mass:
\[
\hat{E}^{+,\mathrm{compound}}_{ug} = \frac{E^{+,\mathrm{compound}}_{ug}}{\sum_h E^{+,\mathrm{compound}}_{uh} + \epsilon},
\]
\[
\hat{E}^{+,\mathrm{genetic}}_{ug} = \frac{E^{+,\mathrm{genetic}}_{ug}}{\sum_h E^{+,\mathrm{genetic}}_{uh} + \epsilon}.
\]

Then combine modalities with weights $\lambda_c$ and $\lambda_g$:
\begin{equation}
S^{+}_{ug} = \lambda_c \hat{E}^{+,\mathrm{compound}}_{ug} + \lambda_g \hat{E}^{+,\mathrm{genetic}}_{ug},
\label{eq:morph_positive_score}
\end{equation}
where by default $\lambda_c = \lambda_g = 1/2$ if both are available, and $\lambda=1$ for the available modality if only one exists.

An analogous score $S^-_{ug}$ is constructed from negative similarity evidence $w^-_{ur}$.

\section{Gene program extraction}
\subsection{Program families}
Each query unit $u$ yields one or more gene programs:
\begin{itemize}[leftmargin=2em]
\item \texttt{similar}: based on $S^+_{ug}$,
\item \texttt{opposite}: based on $S^-_{ug}$ if requested.
\end{itemize}
The default is to emit only the similar program. Beyond polarity, the current implementation supports three retrieval objectives:
\begin{itemize}[leftmargin=2em]
\item \texttt{direct\_target}: strict target-first pooling before hard reference truncation,
\item \texttt{mechanism}: family/mechanism-aware reranking and local expansion from the full retained neighborhood,
\item \texttt{hybrid}: emit both a strict core output and a broader expanded output from an independent mechanism branch, while marking one as preferred for downstream use.
\end{itemize}

\subsection{Selection}
Given a nonnegative gene score vector $S_{ug}$ for a chosen polarity, select genes by:
\begin{itemize}[leftmargin=2em]
\item \textbf{top-k} (default): choose the $K$ highest-scoring genes,
\item \textbf{quantile}: choose genes above a quantile threshold,
\item \textbf{threshold}: choose genes above a minimum score.
\end{itemize}
As in other extractors, the implementation should enforce default size bounds of 100--500 genes for GMT emission and warn when fewer genes are available.

\subsection{Neighbor restriction for specificity}
In practice, morphology retrieval can become too diffuse if all weakly similar references are allowed to contribute. A robust default is therefore to retain only the top $K_{\mathrm{nbr}}$ reference neighbors per query and polarity after thresholding and hubness-aware evidence reweighting:
\[
\mathcal{N}_u^+ = \operatorname{TopK}\{r : \mathrm{sim}(u,r) \ge \tau\},
\]
and analogously for the opposite-polarity case using $-\mathrm{sim}(u,r)$.
Only retained neighbors contribute to $E_{ug}$. This improves specificity by preventing a long tail of weak matches from dominating the routed gene evidence. In the practical implementation, hubness-adjusted evidence is computed first and the retained neighborhood is then truncated to a narrow local panel.

\subsection{Target-first aggregation}
For direct-target retrieval, the practical implementation does not immediately map query $\rightarrow$ top references $\rightarrow$ genes. Instead it first forms a positive evidence pool over references, aggregates that evidence by routed target gene, and only then restricts the retained references to those supporting the strongest target candidates. If $T_{ug}$ denotes pooled target support, direct-target mode defines the strict nomination pool from the raw $T_{ug}$ values together with support-count and best-similarity guards, and then uses recurrence-adjusted support only to rank genes within that allowed pool before the final reference cap is applied. This is designed to rescue weak distributed same-target support that would otherwise be lost when one off-target reference has the single highest similarity while still preventing recurrent generic targets from monopolizing the strict branch.

\subsection{Mechanism-aware expansion}
If the bundle includes optional target annotations such as \texttt{target\_family}, \texttt{target\_class}, \texttt{mechanism\_label}, or \texttt{pathway\_seed}, mechanism mode estimates label support directly from the retained mechanism neighborhood rather than from already-reranked core genes. When exact target support is weak but neighborhood support is coherent at a narrower label level, the expanded output boosts genes carrying that supported label. This yields broader, enrichment-ready outputs without pretending that exact-target recovery is stronger than it is.

If these annotations are absent, mechanism mode remains runnable but should be interpreted as a broad target-routing fallback rather than as full family-aware inference. In that setting the implementation should warn explicitly so users do not mistake missing annotation support for a negative biological result.

In the current practical implementation, a canonical curated annotation table ships with the package and is injected into newly built morphology bundles by default unless the user overrides it. The controlled vocabulary has four fields:
\texttt{target\_family}, \texttt{target\_class}, \texttt{mechanism\_label}, and \texttt{pathway\_seed}. The design is intentionally conservative: some recurrent generic genes are left blank on purpose so they do not seed broad family/mechanism expansion from weak morphology neighborhoods.

\subsection{Expansion gating}
Family/mechanism expansion is no longer controlled by a hard veto requiring exact agreement between raw and penalized neighborhoods. Instead, the implementation computes a soft expansion confidence from the retained mechanism neighborhood and treats raw-versus-retained disagreement as a penalty term and diagnostic.

Let $L_u^{\mathrm{pen}}(h,\ell)$ denote retained-neighborhood support for label $\ell$ at hierarchy level $h \in \{\texttt{pathway\_seed}, \texttt{target\_class}, \texttt{mechanism\_label}, \texttt{target\_family}\}$. The practical implementation searches these levels from most specific to least specific and chooses the first level whose top label has sufficient retained support, multiplicity of supporting neighbors, and acceptable modality structure. This gives a chosen level $h^\star$ and label $\ell^\star$ together with an expansion confidence score in $[0,1]$.

Raw-neighborhood summaries are still computed, but only to downweight confidence or surface warnings when they disagree with the retained neighborhood. They are not a binary veto.

The confidence rule is modality-aware:
\begin{itemize}[leftmargin=2em]
\item for \texttt{orf} and \texttt{crispr} queries, same-modality support still contributes strongly to whether a label is trusted for expansion,
\item for \texttt{compound} queries, same-modality support is \emph{not} required, because biologically meaningful compound interpretation often depends on coherent genetic neighbors.
\end{itemize}

This asymmetry is intentional. It preserves ORF/CRISPR specificity while still allowing compound queries to expand through stable genetically supported class or mechanism structure.

\subsection{Branch structure}
The current implementation maintains two distinct routed branches:
\begin{itemize}[leftmargin=2em]
\item a \textbf{core branch}, built from references that support the strongest nominated targets and intended to remain strict,
\item a \textbf{mechanism branch}, built from the full retained coherent neighborhood and used for label choice plus local expansion.
\end{itemize}

In \texttt{hybrid} mode both branches are always computed independently. The strict branch writes \texttt{geneset.core.tsv}; the mechanism branch writes \texttt{geneset.expanded.tsv}; and the public \texttt{geneset.tsv} is formed by merging the two so that core genes are preserved while expanded-only genes add context. Internal preferred-variant diagnostics are still recorded, but the user-facing hybrid output is no longer a winner-take-all copy of the mechanism branch.

\subsection{Local weighted expansion}
Expansion is not implemented as a flat union over all genes carrying any broad matching annotation. Once $(h^\star, \ell^\star)$ has been chosen, the candidate gene universe is restricted in the following order:
\begin{itemize}[leftmargin=2em]
\item genes present in the full pre-exclusion bundle and carrying $(h^\star, \ell^\star)$,
\item if empty, genes observed in the local retained neighborhood carrying $(h^\star, \ell^\star)$,
\item only as a fallback, all genes in the packaged annotation table carrying $(h^\star, \ell^\star)$.
\end{itemize}

For narrow levels such as \texttt{pathway\_seed}, \texttt{target\_class}, and \texttt{mechanism\_label}, a controlled global annotation fallback is allowed even when the effective held-out retrieval panel is sparse. Local evidence still comes only from effective retrieved neighbors, but the bundle prior counts and candidate universe come from the full bundle so held-out nominal targets are not silently removed from consideration. Candidate genes are then weighted by local routed support when available, with only a small bundle-coverage prior for genes that match the label but have weak direct local evidence. This avoids the earlier broad family-union behavior and preserves the preference for the narrowest stable label.

\subsection{Compound promiscuity and optional target confidence}
Compound references can be more weakly informative than genetic references because of polypharmacology and uneven target annotation confidence. The current implementation therefore uses a shared bounded per-reference mapping for compounds: positive target weights are retained, but if their total mass exceeds $1.0$ they are renormalized so that a single compound contributes at most one reference-equivalent unit of routed target mass. This bounded mapping is used consistently in exact-target nomination, label voting, and mechanism gene accumulation, so a multitarget compound can support several genes or labels without voting several full times for the same biological explanation. In addition, if the compound-target table carries optional columns such as \texttt{confidence}, \texttt{primary\_target}, \texttt{potency\_bucket}, or \texttt{source\_confidence}, these are folded into edge weights through a bounded multiplier. Inputs that omit these columns retain the previous behavior.

\subsection{Same-modality-first retrieval}
If the query metadata identifies the perturbation modality (for example compound versus ORF), same-modality neighbors can be prioritized before cross-modality fusion. Cross-modality neighbors are still useful, but they should contribute most strongly when they reinforce the same local target signal as the same-modality neighborhood rather than dominating it outright.

\subsection{Adaptive neighborhood}
In practice, $K_{\mathrm{nbr}}$ is better interpreted as an upper bound than as a fixed neighborhood size. If evidence drops sharply after a small coherent local panel has been assembled, the extractor may stop early rather than adding a long tail of weak or isolated neighbors. This improves specificity and makes the effective retained-neighbor count query-dependent.

\subsection{Within-set normalization}
For the selected set $\mathcal{S}_u$, define within-set weights:
\[
W_{ug} =
\begin{cases}
\displaystyle \frac{S_{ug}}{\sum_{h \in \mathcal{S}_u} S_{uh} + \epsilon} & g \in \mathcal{S}_u, \\
0 & \text{otherwise.}
\end{cases}
\]
This yields a probability-like distribution over selected genes, suitable for weighted enrichment tools.

\section{Reference bundles}
A morphology extractor is most useful when paired with a reusable reference bundle. The recommended v1 bundle pattern is:
\begin{itemize}[leftmargin=2em]
\item \texttt{reference\_profiles.parquet}: consensus perturbation profiles (one row per perturbation/context),
\item \texttt{reference\_metadata.tsv.gz}: perturbation type, cell line, timepoint, QC, control labels,
\item \texttt{compound\_targets.tsv.gz}: compound-to-gene target map with weights,
\item optional \texttt{target\_annotations.tsv.gz}: gene-level \texttt{target\_family}, \texttt{target\_class}, \texttt{mechanism\_label}, and \texttt{pathway\_seed} annotations,
\item \texttt{feature\_schema.tsv.gz}: ordered feature list,
\item \texttt{feature\_stats.tsv.gz}: per-feature center and scale,
\item \texttt{bundle.json}: version, provenance, source URLs, checksums.
\end{itemize}

A practical first bundle should be derived from the JUMP pilot data in a single cell line and timepoint (for example, U2OS 48h or A549 48h), because those datasets contain matched compounds, CRISPR knockouts, and ORF overexpression perturbations with strong metadata and manageable scale compared to the full JUMP production release. \cite{jump_gallery,jump_pilot,jump_target}

Distributed morphology bundles should include \texttt{target\_annotations.tsv.gz} by default. Annotation-free bundles remain useful as strict baselines, but they are not the preferred public distribution because they limit mechanism-aware retrieval.

The practical default should prefer coherent same-timepoint bundles. If a public context does not contain all modalities at one timepoint, the workflow should keep a coherent same-timepoint reference from the available modalities and warn explicitly about missing modalities. Mixed-timepoint bundles should require explicit user intent.

\subsection{Reference hubness penalty}
Some reference perturbations are generic morphology hubs: they are moderately similar to many unrelated queries and can dominate routed gene evidence. Let $h_r$ denote a precomputed hubness score for reference perturbation $r$, estimated from within-bundle reference-reference similarities. A default penalty rescales evidence before routing:
\[
\tilde{w}_{ur} = \frac{w_{ur}}{\epsilon + h_r}.
\]
The implementation may also use a rank-based penalty. In practice this is often more stable than a raw inverse-linear rescaling when hub scores are compressed. The intent is the same: reduce the influence of generic hub-like references and increase specificity of routed gene evidence.

\subsection{Gene recurrence penalty}
Some genes recur across many unrelated morphology perturbations because they are common library targets or generic downstream responders. A weak default penalty can be defined from reference-level recurrence:
\[
\omega_g = \frac{\log((N+1)/(d_g+1))}{\log(N+1)},
\]
where $d_g$ is the number of reference perturbations that route to gene $g$ and $N$ is the total number of mapped references. The final routed gene score can then be rescaled as
\[
\tilde{S}_{ug} = \omega_g S_{ug},
\]
which reduces recurrent generic genes without imposing a hard blacklist.

\subsection{Control-derived calibration}
Control profiles should remain excluded from the target-neighbor candidate set, but they can still be useful for calibration. A simple practical default is to estimate a control-derived nuisance mean in feature space and subtract it from both query and reference vectors before similarity scoring. This aims to reduce broad plate or vehicle-like morphology axes without turning controls into mechanistic neighbors.

A stronger extension is to estimate a low-rank nuisance subspace from controls and residualize against it. Let $\mu^{(C)} \in \mathbb{R}^{F}$ denote the mean control profile and let $U = [u_1,\dots,u_K]$ be the top $K$ orthonormal nuisance axes estimated from centered control profiles. The residualized query and reference vectors are then
\[
\hat{X}^{(Q)}_u = \left(\bar{X}^{(Q)}_u - \mu^{(C)}\right) - \sum_{k=1}^{K} \left\langle \bar{X}^{(Q)}_u - \mu^{(C)}, u_k \right\rangle u_k,
\]
\[
\hat{X}^{(R)}_r = \left(\tilde{X}^{(R)}_r - \mu^{(C)}\right) - \sum_{k=1}^{K} \left\langle \tilde{X}^{(R)}_r - \mu^{(C)}, u_k \right\rangle u_k.
\]
Similarity is then computed on $\hat{X}^{(Q)}$ and $\hat{X}^{(R)}$ rather than on the uncalibrated vectors. In practical CLI terms, this corresponds to a control-residualization mode that falls back to mean-centering when too few controls are available to estimate stable nuisance axes. The point of this extension is to remove a broad nuisance direction even when the control mean itself is near zero.

\section{QC and failure modes}
Important warnings to implement:
\begin{itemize}[leftmargin=2em]
\item Low feature overlap between query and reference.
\item Query profiles dominated by metadata columns or non-numeric fields.
\item Reference control profiles accidentally included as retrieval candidates.
\item Low perturbation QC or very low number of usable reference matches.
\item Gene families dominating selected sets (for example, a broad receptor family), reported as warnings but not filtered by default.
\item Low retrieval confidence, for example when few positive neighbors are retained or the negative-similarity fraction is high.
\item Low specificity confidence, for example when retained neighbors are geometrically coherent but routed gene evidence remains diffuse across many unrelated genes.
\item High retrieval confidence with low specificity confidence, which indicates a stable morphology neighborhood that still does not concentrate on target-like routed gene evidence.
\item Stable raw neighborhood but unstable family/mechanism support after penalties, which indicates that expansion would be too brittle to trust.
\end{itemize}

In practice, the implementation can use specificity confidence as a guardrail rather than a passive annotation. In particular, opposite-polarity programs remain experimental and may be suppressed by default unless routed gene evidence reaches at least a moderate specificity level.

The implementation should also expose routing diagnostics in machine-readable summaries, including:
\begin{itemize}[leftmargin=2em]
\item top raw candidate neighbors,
\item retained neighbors after penalization/filtering,
\item branch-specific retained neighbors for the core and mechanism branches,
\item family vote summaries before and after filtering,
\item mechanism vote summaries before and after filtering,
\item hierarchy-level label support summaries,
\item modality-separated label support summaries,
\item top routed target candidates together with raw versus recurrence-adjusted support,
\item whether compound edges were renormalized and how many references were affected,
\item and an explicit expansion-decision record describing why expansion was allowed or suppressed, which label level was chosen, and how large the local expansion universe was.
\end{itemize}

\section{Recommended defaults}
A sensible initial default is:
\begin{itemize}[leftmargin=2em]
\item reference bundle: JUMP pilot, matched cell line and timepoint,
\item query aggregation: median within user-supplied group,
\item similarity metric: cosine,
\item polarity: similar only,
\item neighbor restriction: top 20 hubness-adjusted reference neighbors with nonnegative similarity retained by default,
\item same-modality-first retrieval when query modality metadata is available,
\item adaptive neighborhood truncation after a minimum coherent local panel,
\item control calibration: mean-centering by default, with control residualization available when enough controls are present,
\item hubness penalty: rank-based downweighting when the bundle provides hub scores,
\item gene recurrence penalty: weak IDF-style downweighting of genes that recur across many reference perturbations,
\item modality balance: equal across available modalities,
\item selection: top-k with $K=200$,
\item normalization: within-set $\ell_1$,
\item GMT emission with 100--500 size bounds.
\end{itemize}

Opposite-polarity programs remain useful as a secondary exploratory view, but should be treated as experimental unless retrieval confidence and context matching are strong.

\begin{thebibliography}{9}

\bibitem{jump_main}
JUMP-Cell Painting Consortium.
JUMP-Cell Painting Consortium public portal.
\url{https://jump-cellpainting.broadinstitute.org/}

\bibitem{jump_gallery}
Cell Painting Gallery documentation: complete datasets.
\url{https://broadinstitute.github.io/cellpainting-gallery/complete_datasets.html}

\bibitem{jump_target}
JUMP-Target metadata repository.
\url{https://github.com/jump-cellpainting/JUMP-Target}

\bibitem{jump_pilot}
Chandrasekaran et al.
Three million images and morphological profiles of cells treated with matched chemical and genetic perturbations.
Nature Methods (2024).

\bibitem{jump_morphmap}
Chandrasekaran et al.
Morphological map of under- and overexpression of genes in human cells.
Nature Methods (2025).

\end{thebibliography}

\end{document}
