\documentclass[11pt]{article}

\usepackage{amsmath, amssymb}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}

\geometry{margin=1in}

\title{Supplementary Note: Alternative Splicing Event Extractors to Gene Programs in dig-gene-set-extractors}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Overview and goals}
This note describes the alternative splicing assay family in \texttt{dig-gene-set-extractors}. The design follows the same repo pattern used elsewhere in the repository: a direct converter for already-scored rows, a matrix front-end for public sample-by-feature matrices, and an optional compact reference bundle that adds harmonization and conservative nuisance priors.

The entrypoints are:
\begin{itemize}[leftmargin=2em]
\item \texttt{splice\_event\_diff}: event-level differential splicing table to signed gene program and GMT gene sets.
\item \texttt{splice\_event\_matrix}: PSI matrix plus sample metadata to one or more contrast-specific gene programs.
\item \texttt{workflows splice\_prepare\_public}: narrow public-data normalizer centered on TCGA SpliceSeq-like PSI matrices.
\item \texttt{workflows splice\_prepare\_reference\_bundle}: builder for a lightweight alias, ubiquity, impact, burden, and dataset-stratified nuisance bundle.
\end{itemize}

This family is intentionally event-centric rather than transcript-centric. LeafCutter, MAJIQ, Whippet, and TCGA SpliceSeq differ in file layout and event definition, but all can be normalized to a common abstraction:
\begin{enumerate}[leftmargin=2em]
\item an event identity together with an identity-confidence tier and event-key namespace,
\item a signed effect such as PSI change or a signed statistic,
\item optional uncertainty information such as p-value, posterior probability, or read support,
\item an event-to-gene mapping,
\item optional conservative event priors,
\item a gene-level aggregation rule that does not reward genes merely for having many measurable rows.
\end{enumerate}

The central modeling problem is therefore not transcript reconstruction from raw reads. It is the conversion of heterogeneous event outputs into a stable, gene-level mechanistic program that is suitable for enrichment analysis and downstream interpretation.

\section{Why event-to-gene projection is the right first extractor target}
Splicing assays do not start from genes. Their natural units are junctions, local splicing variations, cassette exons, retained introns, or broader event groups. A practical extractor should therefore:
\begin{itemize}[leftmargin=2em]
\item harmonize raw event identities whenever possible,
\item score each event with a signed effect weighted by uncertainty,
\item use priors conservatively and in a source-aware way,
\item collapse event scores to genes with protection against duplicate bookkeeping rows, local event stacking, and locus-density artifacts.
\end{itemize}

This is closely analogous to the PTM design already present in the repository, except the relevant nuisance structure is event multiplicity, study leakage, and local genomic concentration rather than site ubiquity alone.

\section{Mental model}
Let $i \in \mathcal{E}$ index splice events and let $s \in \mathcal{S}$ index samples.

For a PSI matrix input, the basic observation is
\[
\psi_{is} \in [0,1],
\]
where $\psi_{is}$ is the percent spliced in for event $i$ in sample $s$.

For a differential table input, common fields include:
\begin{itemize}[leftmargin=2em]
\item an effect size such as $\Delta \psi_i$,
\item a signed statistic $t_i$ or $z_i$,
\item a p-value or adjusted p-value $p_i$ or $q_i$,
\item a posterior probability or confidence score $\pi_i$,
\item support counts or denominators $r_i$,
\item event metadata including type and optional coordinates,
\item one or more mapped genes.
\end{itemize}

The extractor converts event-level evidence into a signed event score $z_i$, aggregates those event scores to a gene score $s_g$, and then converts magnitudes into nonnegative weights $w_g$ for downstream gene-set use.

\section{How CLI arguments map to the model}
\begin{itemize}[leftmargin=2em]
\item Input event table or PSI matrix:
  \begin{itemize}
  \item \texttt{--splice\_tsv} for \texttt{splice\_event\_diff}
  \item \texttt{--psi\_matrix\_tsv}, \texttt{--sample\_metadata\_tsv}, and optional \texttt{--event\_metadata\_tsv} for \texttt{splice\_event\_matrix}
  \end{itemize}
\item Event identity and metadata columns:
  \texttt{--event\_id\_column}, \texttt{--event\_group\_column}, \texttt{--event\_type\_column},
  \texttt{--chrom\_column}, \texttt{--start\_column}, \texttt{--end\_column}, \texttt{--strand\_column},
  \texttt{--gene\_id\_column}, \texttt{--gene\_symbol\_column}.
\item Base event score:
  \texttt{--score\_mode} with supporting columns such as \texttt{--delta\_psi\_column}, \texttt{--stat\_column}, \texttt{--pvalue\_column}, \texttt{--padj\_column}, \texttt{--probability\_column}, or \texttt{--score\_column}.
\item Confidence weighting:
  \texttt{--confidence\_weight\_mode}, \texttt{--read\_support\_column}, \texttt{--min\_read\_support}.
\item Impact priors:
  \texttt{--impact\_mode}, \texttt{--event\_impact\_resource\_id}.
\item Duplicate collapse and event-to-gene mapping:
  \texttt{--event\_dup\_policy}, \texttt{--ambiguous\_gene\_policy}.
\item Gene aggregation and stability controls:
  \texttt{--gene\_aggregation}, \texttt{--gene\_topk\_events},
  \texttt{--gene\_burden\_penalty\_mode}, \texttt{--gene\_support\_penalty\_mode}.
\item Source-aware bundle use:
  \texttt{--resources\_dir}, \texttt{--resources\_manifest}, \texttt{--resource\_policy}, \texttt{--use\_reference\_bundle},
  \texttt{--source\_dataset}, \texttt{--bundle\_same\_dataset\_policy},
  and resource ids for alias, ubiquity, impact, gene burden, dataset-stratified nuisance summaries, and an optional gene-locus table.
\item Locality controls:
  \texttt{--locus\_density\_penalty\_mode}, \texttt{--locus\_density\_window\_bp}, \texttt{--locus\_density\_top\_n}.
\item Gene selection and weighting:
  \texttt{--select}, \texttt{--top\_k}, \texttt{--quantile}, \texttt{--min\_score}, \texttt{--normalize}.
\item GMT export:
  \texttt{--emit\_gmt}, \texttt{--gmt\_split\_signed}, \texttt{--gmt\_topk\_list}, \texttt{--gmt\_mass\_list}.
\end{itemize}

\section{Formal model}
Let $i \in \mathcal{E}$ index canonicalized event records and let $g \in \mathcal{G}$ index genes. The pipeline has six logical layers:
\begin{enumerate}[leftmargin=2em]
\item assign a canonical event identity, confidence tier, and namespace,
\item define a signed base event score $x_i$,
\item apply uncertainty and prior weights to obtain a final event score $z_i$,
\item collapse duplicate rows and local event-group dependence,
\item aggregate events to gene scores with conservative burden and support shrinkage,
\item optionally apply locus-diversity suppression, then select and normalize the gene program.
\end{enumerate}

\subsection{Step 1: event identity, confidence tier, and namespace}
Each raw row is mapped to a canonical event key $k(i)$, a confidence tier $c(i)$, and a namespace $n(i)$.

Preferred order:
\begin{enumerate}[leftmargin=2em]
\item If an alias bundle maps a raw event key to a shared key, use the bundle key.
\item Otherwise, if stable coordinates and event type are available, construct a deterministic coordinate-based key.
\item Otherwise, if the source family is TCGA SpliceSeq-like and the row exposes a structured AS identifier together with gene and event type, construct a source-family-stable key.
\item Otherwise, fall back to a sanitized raw identifier.
\end{enumerate}

A coordinate-based key can use a tuple such as
\[
(\mathrm{chrom}, \mathrm{start}, \mathrm{end}, \mathrm{strand}, \mathrm{event\_type}, \mathrm{event\_group}).
\]

The implemented identity tiers are:
\[
(c(i), n(i)) =
\begin{cases}
(\texttt{high}, \texttt{global\_coordinate}) & \text{coordinate-canonical event},\\
(\texttt{medium}, \texttt{tcga\_spliceseq\_asid}) & \text{TCGA SpliceSeq source-family-stable key},\\
(\texttt{low}, \texttt{raw\_id\_fallback}) & \text{generic raw fallback.}
\end{cases}
\]

The medium tier is intentionally narrow. It is only for TCGA SpliceSeq-style event identities and should be treated as reusable within that source family, not as a general cross-tool splice identity.

\subsection{Step 2: base event score}
The extractor supports several base score definitions.

\paragraph{A1. \texttt{score\_mode=auto} (default).}
Resolve in the following order:
\begin{itemize}[leftmargin=2em]
\item \texttt{stat}, if a signed statistic is available,
\item \texttt{delta\_psi\_times\_neglog10p}, if $\Delta \psi_i$ and p-value information are available,
\item \texttt{delta\_psi\_times\_confidence}, if $\Delta \psi_i$ and posterior or support fields are available,
\item \texttt{custom\_column}, if an explicit score column was requested.
\end{itemize}

For \texttt{splice\_event\_matrix}, per-event p-values are Benjamini-Hochberg adjusted within each emitted contrast before the shared scorer runs, so \texttt{score\_mode=auto} preferentially uses \texttt{padj} when it can be computed.

\paragraph{A2. \texttt{score\_mode=stat}.}
If a signed statistic $t_i$ is available,
\[
x_i = t_i.
\]

\paragraph{A3. \texttt{score\_mode=delta\_psi\_times\_neglog10p}.}
If a signed PSI effect $\Delta \psi_i$ and p-value $q_i$ are available,
\begin{equation}
x_i = \Delta \psi_i \cdot \min\{-\log_{10}(q_i + \epsilon), C\},
\label{eq:dpsi_p}
\end{equation}
with small $\epsilon > 0$ and cap $C$.

\paragraph{A4. \texttt{score\_mode=delta\_psi\_times\_confidence}.}
If a signed PSI effect $\Delta \psi_i$ and posterior or confidence $\pi_i$ are available,
\begin{equation}
x_i = \Delta \psi_i \cdot \pi_i.
\label{eq:dpsi_conf}
\end{equation}

\paragraph{A5. \texttt{score\_mode=custom\_column}.}
If the user provides a custom signed score column,
\[
x_i = \mathrm{custom}(i).
\]

\subsection{Step 3: confidence weighting}
Differential splicing outputs are noisy, so the base event score is shrunk by bounded uncertainty weights.

Define optional components:
\begin{align}
w_i^{(p)} &= \frac{\min\{-\log_{10}(p_i + \epsilon), C\}}{C}, \\
w_i^{(\pi)} &= \mathrm{clip}\!\left(\frac{\pi_i - \pi_0}{1-\pi_0}, 0, 1\right), \\
w_i^{(r)} &= \min\!\left\{\frac{\log(1+r_i)}{\log(1+r_0)}, 1\right\}.
\end{align}

Then
\[
\omega_i =
\begin{cases}
1 & \texttt{confidence\_weight\_mode=none},\\
w_i^{(p)} & \texttt{confidence\_weight\_mode=pvalue},\\
w_i^{(\pi)} & \texttt{confidence\_weight\_mode=probability},\\
w_i^{(r)} & \texttt{confidence\_weight\_mode=read\_support},\\
w_i^{(p)} w_i^{(\pi)} w_i^{(r)} & \texttt{confidence\_weight\_mode=combined}.
\end{cases}
\]

This weighting can only shrink weak evidence. It does not reverse sign and is not allowed to dominate the base score.

\subsection{Step 4: conservative impact prior}
The extractor can optionally use a compact impact bundle. This remains conservative in v1 and does not attempt full transcript-consequence inference.

Let $b_i^{(\mathrm{coding})}$, $b_i^{(\mathrm{frame})}$, $b_i^{(\mathrm{nmd})}$, and $b_i^{(\mathrm{novel})}$ be bounded indicators from event annotations or the bundle. Define a raw impact multiplier
\begin{equation}
\widetilde{\eta}_i =
\exp\!\left(
\beta_{\mathrm{coding}} b_i^{(\mathrm{coding})}
+ \beta_{\mathrm{frame}} b_i^{(\mathrm{frame})}
+ \beta_{\mathrm{nmd}} b_i^{(\mathrm{nmd})}
- \beta_{\mathrm{novel}} b_i^{(\mathrm{novel})}
\right),
\label{eq:impact_raw}
\end{equation}
with small coefficients so that $\widetilde{\eta}_i$ stays close to $1$.

Let $\rho_i \in [0,1]$ be an evidence-quality term. The usable impact prior is shrunk toward neutrality:
\begin{equation}
\eta_i = \mathrm{clip}\!\left(1 + \rho_i(\widetilde{\eta}_i - 1), \eta_{\min}, \eta_{\max}\right).
\label{eq:impact_shrunk}
\end{equation}

Recommended defaults are narrow, for example $\eta_{\min}=0.75$ and $\eta_{\max}=1.35$.

Impact use is stricter for non-high-confidence identities:
\begin{itemize}[leftmargin=2em]
\item high-confidence events can use the bundle impact prior directly, after bounded shrinkage,
\item medium-confidence events can use impact priors only within their matching namespace,
\item low-confidence raw fallbacks are neutral unless the bundle reports recurrence across at least two non-self datasets.
\end{itemize}

\subsection{Step 5: ubiquity prior and same-dataset exclusion}
Frequently observed events should be somewhat less informative than rarer events. The bundle therefore stores a pooled detection-frequency prior together with dataset-stratified nuisance tables.

Let $N$ be the number of reference samples and let $df_i$ be the number of reference samples where event $i$ is observed. A pooled prior can be written as
\begin{equation}
u_i = \log\!\left(1 + \frac{N+\tau}{df_i+\tau}\right).
\label{eq:idf}
\end{equation}

When the target run knows its own source dataset $d^\star$ and the bundle contains dataset-level nuisance summaries, the runtime first removes matching same-dataset contributions under the default \texttt{bundle\_same\_dataset\_policy=exclude}. If no non-self reference remains, the prior becomes neutral:
\[
u_i^{(-d^\star)} = 1.
\]

If non-self references remain, the runtime replaces the pooled event prior with an effective leave-one-dataset-out prior based on remaining dataset counts:
\[
u_i^{(-d^\star)} \approx \log\!\left(1 + \frac{D_{\neg d^\star}+1}{df_i^{(\mathrm{dataset}, \neg d^\star)}+1}\right),
\]
where $D_{\neg d^\star}$ is the number of remaining reference datasets and $df_i^{(\mathrm{dataset}, \neg d^\star)}$ is the number of remaining datasets in which event $i$ appears.

The final usable ubiquity prior is tier- and namespace-aware:
\[
u_i^\star =
\begin{cases}
u_i^{(-d^\star)} & c(i)=\texttt{high},\\
1 + \lambda_{\mathrm{med}} \left(u_i^{(-d^\star)} - 1\right) & c(i)=\texttt{medium} \text{ and } n(i)=n_{\mathrm{prior}}(i),\\
1 & c(i)=\texttt{medium} \text{ and } n(i)\neq n_{\mathrm{prior}}(i),\\
1 & c(i)=\texttt{low},
\end{cases}
\]
with a moderate shrinkage factor such as $\lambda_{\mathrm{med}} \approx 0.6$.

So the runtime behavior is:
\begin{itemize}[leftmargin=2em]
\item \texttt{high}: strongest prior use,
\item \texttt{medium}: bounded prior use only within the matching source-family namespace,
\item \texttt{low}: event-to-gene mapping can still benefit from aliasing, but ubiquity is neutralized.
\end{itemize}

\subsection{Step 6: final event score}
The final event score is
\begin{equation}
z_i = T(x_i) \cdot \phi_i \cdot \omega_i \cdot \eta_i \cdot u_i^\star,
\label{eq:final_event_score}
\end{equation}
where the score transform is
\[
T(x) =
\begin{cases}
x & \texttt{signed},\\
|x| & \texttt{abs},\\
\max(x,0) & \texttt{positive},\\
\max(-x,0) & \texttt{negative}.
\end{cases}
\]

Here $\phi_i$ is an optional soft effect-size floor when a usable $\Delta \psi_i$ is available:
\[
\phi_i =
\begin{cases}
\min\{1, |\Delta \psi_i| / \delta_0\} & \text{if a usable } \Delta \psi_i \text{ exists},\\
1 & \text{otherwise,}
\end{cases}
\]
with a conservative default such as $\delta_0 = 0.05$.

\section{Duplicate-row collapse}
Multiple raw rows can map to the same canonical event key. Supported policies are:
\begin{itemize}[leftmargin=2em]
\item \texttt{highest\_confidence} (recommended default): retain the row with the largest confidence proxy, breaking ties by largest $|z_i|$,
\item \texttt{max\_abs}: keep the row with the largest $|z_i|$,
\item \texttt{mean}: average signed scores across duplicates,
\item \texttt{sum}: sum signed scores across duplicates.
\end{itemize}

The default favors one representative row rather than inflating genes with repeated bookkeeping records.

\section{Event-to-gene mapping}
Let $M_i \subseteq \mathcal{G}$ denote the mapped genes for event $i$.

Supported policies are:
\begin{itemize}[leftmargin=2em]
\item \texttt{drop} (recommended default): discard ambiguous multi-gene events,
\item \texttt{split\_equal}: distribute $z_i / |M_i|$ to each mapped gene,
\item \texttt{first}: choose the first mapped gene after deterministic sorting.
\end{itemize}

The default remains conservative because ambiguous event-to-gene mappings can create false biological narratives.

\section{Gene aggregation}
Let $E_g$ be the set of event contributions assigned to gene $g$.

\subsection{G0. Event-group-aware independence collapse}
Several rows from one LeafCutter cluster, MAJIQ LSV, or similar local family are not independent evidence. The runtime therefore defines an event-group key $c_i$ from \texttt{event\_group} when available, falling back to the canonical event key otherwise, and keeps one representative per $(g, c_i)$:
\[
E_g^\star = \{ \text{best representative of each } (g, c_i) \}.
\]
The representative is chosen by strongest absolute final event score, with confidence as a tiebreaker.

\subsection{G1. \texttt{gene\_aggregation=signed\_topk\_mean} (recommended)}
Sort event contributions by decreasing absolute magnitude. Let $\mathrm{TopK}(g)$ be the top $K$ event-group-collapsed contributions for gene $g$. Then
\begin{equation}
s_g = \frac{1}{|\mathrm{TopK}(g)|} \sum_{i \in \mathrm{TopK}(g)} z_i.
\label{eq:topk_gene}
\end{equation}

This default avoids rewarding genes simply for having many measurable rows while still allowing a gene to be supported by more than one coherent event.

\subsection{G1b. Conservative burden shrinkage}
Even after top-$K$ aggregation, some cohorts still favor genes with many measurable events. The runtime therefore applies a conservative multiplicity penalty:
\[
\widetilde{s}_g = s_g \cdot \beta_g,
\qquad
\beta_g = \max\!\left(\beta_{\min}, \sqrt{\frac{K}{\max(K, n_g)}}\right),
\]
where $n_g$ is a burden count for gene $g$ and $\beta_{\min}$ is a conservative lower clip such as $0.35$.

The burden count can come from:
\begin{itemize}[leftmargin=2em]
\item the current input after duplicate and event-group collapse, or
\item a compact reference-bundle table such as \texttt{splice\_gene\_event\_burden\_human\_v1.tsv.gz}.
\end{itemize}

In the reference-bundle path, $n_g$ is derived from cross-study nuisance summaries such as unique event groups per gene, number of supporting studies, and median unique groups per study. When dataset-stratified burden tables are available, same-dataset contributions are excluded before this prior is used. If no non-self burden data remain, the burden prior becomes neutral.

\subsection{G1c. Soft gene-support and sign-coherence shrinkage}
Genes supported by multiple independent event groups with coherent sign should be ranked more stably than genes supported by one weak or internally inconsistent signal. After event-group collapse and base aggregation, the runtime therefore computes:
\begin{align}
n_g^{(\mathrm{grp})} &= \text{number of independent event groups retained for } g,\\
n_g^{(\mathrm{evt})} &= \text{number of retained event contributions used for } g,\\
\kappa_g &= \frac{\left|\sum_{i \in \mathrm{TopK}(g)} z_i\right|}{\sum_{i \in \mathrm{TopK}(g)} |z_i|}.
\end{align}
Here $\kappa_g \in [0,1]$ is a sign-coherence statistic.

Under \texttt{gene\_support\_penalty\_mode=auto}, the runtime applies a modest support factor
\[
\gamma_g = \max\!\left(0.55, f_{\mathrm{grp}}(n_g^{(\mathrm{grp})}) \cdot f_{\mathrm{coh}}(\kappa_g)\right),
\]
with a conservative implementation such as
\[
f_{\mathrm{grp}}(n) =
\begin{cases}
1 & n \ge 2,\\
0.85 & n = 1,
\end{cases}
\qquad
f_{\mathrm{coh}}(\kappa) = 0.75 + 0.25\kappa.
\]

The support-adjusted score is then
\[
\widehat{s}_g = \widetilde{s}_g \cdot \gamma_g.
\]
This is intentionally soft. Genes with a single retained group are not dropped; they are only downweighted modestly relative to equally strong genes with better support structure.

\subsection{G2. Alternatives}
Supported alternatives remain:
\begin{itemize}[leftmargin=2em]
\item \texttt{max\_abs},
\item \texttt{sum},
\item \texttt{mean}.
\end{itemize}

For splicing, \texttt{sum} is often too sensitive to event multiplicity and is not recommended as the default.

\section{Active locus-diversity suppression}
Some cohort-scale runs produce top genes that are implausibly concentrated within one genomic window, chromosome arm, or whole chromosome. The extractor therefore supports both diagnostic and active locus-density handling.

Let genes be ordered by decreasing $|\widehat{s}_g|$. Using either event-derived coordinates or an optional compact gene-locus table, define concentration summaries among the top $N$ genes:
\[
f_{\mathrm{window}}, \qquad f_{\mathrm{arm}}, \qquad f_{\mathrm{chrom}},
\]
where each quantity is the fraction of the top-$N$ genes falling in the dominant local window, chromosome arm, or chromosome.

Under \texttt{locus\_density\_penalty\_mode=auto}, the runtime checks whether one of these concentrations exceeds conservative thresholds. If no likely artifact is detected, no penalty is applied. If a likely artifact is detected, the top local representative is preserved and later genes from the same overrepresented region are softly downweighted. A representative form is
\[
\ell_g =
\begin{cases}
1 & \text{first representative from a local region},\\
\frac{1}{\sqrt{m_g}} & \text{later window- or arm-level duplicate},\\
\frac{1}{1 + 0.3(m_g - 1)} & \text{later chromosome-level duplicate},
\end{cases}
\]
where $m_g$ is the repetition count encountered while traversing the ranked list.

The final score entering selection is
\[
s_g^{(\mathrm{final})} = \widehat{s}_g \cdot \ell_g.
\]

This keeps the strongest local gene unchanged while reducing the chance that an entire program is dominated by one locus. The implementation records diagnostics such as \texttt{top\_n\_fraction\_same\_chromosome}, \texttt{top\_n\_fraction\_same\_chromosome\_arm}, \texttt{top\_n\_fraction\_same\_window}, and \texttt{likely\_locus\_density\_artifact}.

\section{From gene scores to a compact gene program}
After all aggregation and optional locus suppression, define the magnitude
\[
m_g = |s_g^{(\mathrm{final})}|.
\]
Select a subset $S \subseteq \mathcal{G}$ using shared repo-wide strategies such as \texttt{top\_k}, \texttt{threshold}, \texttt{quantile}, or \texttt{none}. Then compute nonnegative weights:
\begin{equation}
w_g =
\begin{cases}
\displaystyle \frac{m_g}{\sum_{h \in S} m_h} & \text{if } g \in S \text{ and } \texttt{normalize=within\_set\_l1},\\
0 & \text{otherwise.}
\end{cases}
\label{eq:weights}
\end{equation}

The signed score and nonnegative weight are both preserved in output tables.

\section{Matrix front-end}
\texttt{splice\_event\_matrix} estimates within-study event-level contrasts from a PSI matrix and then reuses the same downstream scorer used by \texttt{splice\_event\_diff}.

\subsection{Input contract}
Required:
\begin{itemize}[leftmargin=2em]
\item \texttt{psi\_matrix.tsv}: one row per event, one column per sample, plus optional event metadata columns.
\item \texttt{sample\_metadata.tsv}: at least \texttt{sample\_id}; usually \texttt{condition} and possibly \texttt{group}.
\end{itemize}

Optional:
\begin{itemize}[leftmargin=2em]
\item \texttt{event\_metadata.tsv}, if event metadata are not embedded in the PSI matrix.
\item \texttt{coverage\_matrix.tsv}, if a public source supplies event-level support.
\end{itemize}

\subsection{Study contrast modes}
Supported contrast modes are:
\begin{itemize}[leftmargin=2em]
\item \texttt{condition\_a\_vs\_b}
\item \texttt{group\_vs\_rest}
\item \texttt{condition\_within\_group}
\item \texttt{baseline}
\end{itemize}

\subsection{Effect metrics}
Recommended options are:
\begin{itemize}[leftmargin=2em]
\item \texttt{welch\_t} (preferred default)
\item \texttt{mean\_diff}
\end{itemize}

For a two-group contrast with groups $A$ and $B$,
\[
\Delta \psi_i = \overline{\psi}_{iA} - \overline{\psi}_{iB}.
\]
If \texttt{effect\_metric=welch\_t}, the front-end also computes
\[
t_i = \frac{\overline{\psi}_{iA} - \overline{\psi}_{iB}}{\sqrt{\frac{s_{iA}^2}{n_A} + \frac{s_{iB}^2}{n_B}}},
\]
using only samples with finite PSI under the missingness rule. Event-level p-values are BH-adjusted within each emitted contrast before the shared scorer runs.

\section{Public-data normalizer}
The public normalizer is intentionally narrow in v1 and is centered on TCGA SpliceSeq-like data.

\subsection{Recommended workflow}
\texttt{workflows splice\_prepare\_public}:
\begin{enumerate}[leftmargin=2em]
\item ingests TCGA SpliceSeq-like PSI tables and sample annotations,
\item standardizes event ids and metadata,
\item writes
  \begin{itemize}
  \item \texttt{psi\_matrix.tsv}
  \item \texttt{sample\_metadata.tsv}
  \item \texttt{event\_metadata.tsv}
  \item \texttt{sample\_id\_map.tsv}
  \item \texttt{event\_id\_map.tsv}
  \item \texttt{prepare\_summary.json}
  \item \texttt{bundle\_source\_row.tsv}
  \end{itemize}
\item preserves numeric PSI values exactly,
\item records coercions, dropped samples, parser status, and metadata provenance in the summary.
\end{enumerate}

The workflow accepts common public aliases such as \texttt{symbol} for \texttt{gene\_symbol} and \texttt{as\_id} for \texttt{event\_id}. If explicit sample annotations are absent, a narrow fallback can infer tumor versus adjacent-normal from the \texttt{\_Norm} suffix, recording the inference rule and barcode diagnostics in \texttt{prepare\_summary.json}.

For TCGA SpliceSeq-like rows without explicit coordinates, the staging workflow promotes structured AS identifiers to the medium-confidence source-family tier:
\begin{itemize}[leftmargin=2em]
\item \texttt{canonicalization\_status=source\_family\_stable\_id}
\item \texttt{canonicalization\_confidence=medium}
\item \texttt{event\_key\_namespace=tcga\_spliceseq\_asid}
\end{itemize}
This is stronger than a generic raw fallback but still weaker than coordinate-based harmonization.

\section{Reference bundle design}
The v1 splicing bundle is intentionally compact. It does not pool raw PSI distributions across unrelated studies for direct biological scoring. Instead, it provides:
\begin{itemize}[leftmargin=2em]
\item event alias harmonization,
\item event ubiquity priors,
\item conservative event impact priors,
\item a compact per-gene event burden table,
\item dataset-stratified nuisance summaries for same-dataset exclusion,
\item an optional compact gene-locus table for locality-aware suppression.
\end{itemize}

\subsection{Recommended files}
\begin{itemize}[leftmargin=2em]
\item \texttt{splice\_event\_aliases\_human\_v1.tsv.gz}
\item \texttt{splice\_event\_ubiquity\_human\_v1.tsv.gz}
\item \texttt{splice\_event\_ubiquity\_by\_dataset\_human\_v1.tsv.gz}
\item \texttt{splice\_event\_impact\_human\_v1.tsv.gz}
\item \texttt{splice\_gene\_event\_burden\_human\_v1.tsv.gz}
\item \texttt{splice\_gene\_event\_burden\_by\_dataset\_human\_v1.tsv.gz}
\item optional \texttt{splice\_gene\_locus\_human\_v1.tsv.gz}
\item \texttt{bundle\_provenance.json}
\item \texttt{local\_resources\_manifest.json}
\end{itemize}

\subsection{Alias schema}
Required columns:
\begin{itemize}[leftmargin=2em]
\item \texttt{input\_event\_key}
\item \texttt{canonical\_event\_key}
\item \texttt{canonicalization\_status}
\item \texttt{canonicalization\_confidence}
\item \texttt{event\_key\_namespace}
\item \texttt{gene\_id}
\item \texttt{gene\_symbol}
\item \texttt{event\_type}
\item \texttt{chrom}
\item \texttt{start}
\item \texttt{end}
\item \texttt{strand}
\item \texttt{source\_dataset}
\end{itemize}

\subsection{Ubiquity schema}
Required columns:
\begin{itemize}[leftmargin=2em]
\item \texttt{canonical\_event\_key}
\item \texttt{gene\_id}
\item \texttt{gene\_symbol}
\item \texttt{event\_type}
\item \texttt{canonicalization\_status}
\item \texttt{canonicalization\_confidence}
\item \texttt{event\_key\_namespace}
\item \texttt{n\_samples\_ref}
\item \texttt{df\_ref}
\item \texttt{fraction\_ref}
\item \texttt{idf\_ref}
\item \texttt{n\_datasets\_ref}
\item \texttt{fraction\_datasets\_ref}
\end{itemize}

\subsection{Impact schema}
Required columns:
\begin{itemize}[leftmargin=2em]
\item \texttt{canonical\_event\_key}
\item \texttt{gene\_id}
\item \texttt{gene\_symbol}
\item \texttt{event\_type}
\item \texttt{canonicalization\_status}
\item \texttt{canonicalization\_confidence}
\item \texttt{event\_key\_namespace}
\item \texttt{impact\_weight\_raw}
\item \texttt{impact\_evidence}
\item \texttt{annotation\_status}
\item \texttt{n\_datasets\_ref}
\end{itemize}

\subsection{Gene-burden schema}
Required columns:
\begin{itemize}[leftmargin=2em]
\item \texttt{gene\_symbol}
\item \texttt{n\_canonical\_events\_ref}
\item \texttt{n\_high\_confidence\_events\_ref}
\item \texttt{n\_medium\_confidence\_events\_ref}
\item \texttt{n\_low\_confidence\_events\_ref}
\item \texttt{n\_unique\_event\_groups\_ref}
\item \texttt{n\_studies\_ref}
\item \texttt{n\_studies\_high\_confidence\_ref}
\item \texttt{fraction\_low\_confidence\_events\_ref}
\item \texttt{median\_unique\_groups\_per\_study}
\end{itemize}

\subsection{Dataset-stratified nuisance summaries}
The bundle also carries compact dataset-level summaries such as:
\begin{itemize}[leftmargin=2em]
\item \texttt{splice\_event\_ubiquity\_by\_dataset\_human\_v1.tsv.gz}
\item \texttt{splice\_gene\_event\_burden\_by\_dataset\_human\_v1.tsv.gz}
\end{itemize}

These files are not intended for direct biological interpretation. Their role is defensive: at scoring time they let the runtime remove same-dataset contributions so a naively pooled bundle can behave like a leave-one-dataset-out prior when provenance is available.

\subsection{Confidence tiers and namespaces}
The bundle makes identity confidence and namespace explicit:
\begin{itemize}[leftmargin=2em]
\item \texttt{coordinate\_canonical} / \texttt{high} / \texttt{global\_coordinate}
\item \texttt{source\_family\_stable\_id} / \texttt{medium} / \texttt{tcga\_spliceseq\_asid}
\item \texttt{raw\_id\_fallback} / \texttt{low} / \texttt{raw\_id\_fallback}
\end{itemize}

The practical meaning is:
\begin{itemize}[leftmargin=2em]
\item high-confidence coordinate keys are the closest thing this assay family has to cross-study harmonized event identities,
\item medium-confidence TCGA keys are reusable across TCGA SpliceSeq-like studies only,
\item low-confidence raw IDs are cohort-local and should not carry strong cross-study priors.
\end{itemize}

\section{QC and artifact detection}
Splicing programs can be biologically rich, but they are vulnerable to technical and representational artifacts. The extractor therefore emits explicit summaries and warnings.

Recommended summary fields include:
\begin{itemize}[leftmargin=2em]
\item counts of input rows, canonical events, and output genes,
\item counts of independent event groups after event-group collapse,
\item event-type composition,
\item fraction of events matched to alias, ubiquity, and impact priors,
\item canonicalization confidence counts for retained events and selected support events,
\item same-dataset exclusion / neutralization counts for event ubiquity and gene burden,
\item fraction of ambiguous multi-gene events dropped,
\item fraction of selected genes driven by a single event,
\item sign-coherence and independent-group support summaries for selected genes,
\item fraction of low-support events,
\item fraction of novel events,
\item delta-PSI soft-floor downweight counts,
\item counts of low-confidence priors neutralized,
\item counts of medium-confidence namespace mismatches neutralized,
\item genomic concentration among top genes.
\end{itemize}

Recommended warning heuristics include:
\begin{itemize}[leftmargin=2em]
\item \textbf{retained intron dominance},
\item \textbf{novel event dominance},
\item \textbf{event multiplicity bias},
\item \textbf{low confidence},
\item \textbf{bundle same dataset match},
\item \textbf{locus density artifact}.
\end{itemize}

Warnings are written both to stderr and to structured summary outputs.

\section{Recommended defaults}
A reasonable default configuration is:
\begin{itemize}[leftmargin=2em]
\item \texttt{score\_mode=auto}
\item \texttt{score\_transform=signed}
\item \texttt{confidence\_weight\_mode=combined}
\item \texttt{impact\_mode=conservative}
\item \texttt{event\_dup\_policy=highest\_confidence}
\item \texttt{gene\_aggregation=signed\_topk\_mean}
\item \texttt{gene\_topk\_events=3}
\item \texttt{gene\_burden\_penalty\_mode=auto}
\item \texttt{gene\_support\_penalty\_mode=auto}
\item \texttt{locus\_density\_penalty\_mode=none}
\item \texttt{delta\_psi\_soft\_floor=0.05}, \texttt{delta\_psi\_soft\_floor\_mode=auto}
\item \texttt{ambiguous\_gene\_policy=drop}
\item \texttt{use\_reference\_bundle=true}
\item \texttt{resource\_policy=skip}
\item \texttt{bundle\_same\_dataset\_policy=exclude}
\item \texttt{select=top\_k}, \texttt{top\_k=200}
\item \texttt{normalize=within\_set\_l1}
\item \texttt{emit\_gmt=true}, \texttt{gmt\_split\_signed=true}, \texttt{gmt\_topk\_list=200}
\end{itemize}

For matrix inputs:
\begin{itemize}[leftmargin=2em]
\item \texttt{effect\_metric=welch\_t}
\item \texttt{missing\_value\_policy=min\_present}
\item \texttt{min\_present\_per\_condition=3}
\item \texttt{min\_samples\_per\_condition=3}
\end{itemize}

For public TCGA-like cohort runs, \texttt{locus\_density\_penalty\_mode=auto} is often a better operational choice than the conservative global default \texttt{none}, because it can suppress chromosome- or window-dominated artifact programs while preserving the strongest local representative.

\section{What should not be in v1}
The v1 extractor deliberately avoids overreach:
\begin{itemize}[leftmargin=2em]
\item no heavy transcript-level consequence engine,
\item no transcript assembly from raw reads,
\item no aggressive consequence priors without strong shrinkage,
\item no gene aggregation rule that rewards genes merely for having many measurable junctions,
\item no broad promotion of non-TCGA raw event ids into medium confidence without explicit source-family evidence.
\end{itemize}

\section{Interpretation}
This design should produce gene programs that are distinctive when splicing is genuinely mechanistic, for example:
\begin{itemize}[leftmargin=2em]
\item programs driven by spliceosome or RNA-binding protein perturbation,
\item exon-skipping or junction-switching programs that map to coherent pathways,
\item intron-retention programs consistent with stress or altered post-transcriptional control.
\end{itemize}

At the same time, the extractor preserves enough diagnostics to distinguish plausible biology from artifacts. That balance is the main reason to prefer event-to-gene projection plus conservative, source-aware priors over a heavier transcript-consequence stack in the initial implementation.

\section*{References}
\begin{thebibliography}{9}

\bibitem{tcga_spliceseq}
Ryan M, Wong WC, Brown R, et al.
TCGA SpliceSeq a compendium of alternative mRNA splicing in cancer.
\emph{Nucleic Acids Research}. 2016;44(D1):D1018-D1022.

\bibitem{leafcutter}
Li YI, Knowles DA, Pritchard JK.
Annotation-free quantification of RNA splicing using LeafCutter.
\emph{Nature Genetics}. 2018;50(1):151-158.

\bibitem{majiq}
Vaquero-Garcia J, Barrera A, Gazzara MR, et al.
A new view of transcriptome complexity and regulation through the lens of local splicing variations.
\emph{eLife}. 2016;5:e11752.

\bibitem{whippet}
Sterne-Weiler T, Weatheritt RJ, Best AJ, Ha KCH, Blencowe BJ.
Efficient and accurate quantitative profiling of alternative splicing patterns of any complexity on a laptop.
\emph{Molecular Cell}. 2018;72(1):187-200.e6.

\end{thebibliography}

\end{document}
